---
title: "Study Comparison"
author: "Nicolas Banholzer"
date: "2023-07-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(tidyverse)
library(reshape2)

library(rstan)
library(brms)
library(tidybayes)
library(modelr)

source("../utils/epi.r")
source("../utils/bayes.r")
source("../utils/plotting.r")
source("../utils/tex.r")
```


## Data

```{r}
# saliva
saliva_2022 <- readxl::read_xlsx("../data-clean/comparison/mcid-1-molecular-results.xlsx", "Saliva")
saliva_2023 <- readxl::read_xlsx("../data-clean/comparison/mcid-2-molecular-results.xlsx", "Saliva")

saliva <- rbind(saliva_2022 %>% mutate(year_num = 2022),
                saliva_2023 %>% mutate(year_num = 2023))

# bioaerosols
air_2022 <- readxl::read_xlsx("../data-clean/comparison/mcid-1-molecular-results.xlsx", "Bioaerosols")
air_2023 <- readxl::read_xlsx("../data-clean/comparison/mcid-2-molecular-results.xlsx", "Bioaerosols")

air <- rbind(air_2022 %>% mutate(year_num = 2022),
             air_2023 %>% mutate(year_num = 2023))

# viral loads
vl_2022 <- readxl::read_xlsx("../data-clean/comparison/mcid-1-molecular-results.xlsx", "Viral Loads")
vl_2023 <- readxl::read_xlsx("../data-clean/comparison/mcid-2-molecular-results.xlsx", "Viral Loads")

vl <- rbind(vl_2022 %>% mutate(year_num = 2022),
            vl_2023 %>% mutate(year_num = 2023))

# air filters
# TODO: just list

# other data
odf <- readRDS("../data-clean/comparison/merged-data-both-years-all-classes.rds")
```


## Virus distribution

```{r}
saliva_sum <- saliva %>%
  mutate(IF = IFA + IFB) %>%
  dplyr::select(year_num, IF, HRV, AdV, CoV, MPV, PIV) %>%
  melt("year_num") %>%
  rename(virus = variable,
         n = value) %>%
  group_by(year_num, virus) %>%
  summarize(n = sum(n)) %>%
  ungroup() %>%
  filter(n > 0) %>%
  mutate(year = ifelse(year_num == 2022, "2021/22", "2022/23"),
         year = factor(year, levels = c("2021/22", "2022/23")),
         virus = factor(virus, levels = c("IF", "HRV", "AdV", "CoV", "MPV", "PIV"))) %>%
  filter(!is.na(n)) %>%
  group_by(year) %>%
  arrange(virus) %>%
  mutate(p = n / sum(n, na.rm = T),
         ymax = cumsum(p),
         ymin = lag(ymax, n = 1),
         ymin = ifelse(is.na(ymin), 0, ymin),
         labelPosition = (ymax + ymin) / 2,
         label = paste0(virus, ": ", n)) %>%
  ungroup() %>%
  mutate(labelPosition = ifelse(year == "2022/23" & virus == "MPV", labelPosition - .015, labelPosition),
         labelPosition = ifelse(year == "2021/22" & virus == "IF", ymin, labelPosition),
         labelPosition = ifelse(year == "2021/22" & virus == "AdV", ymax, labelPosition))

saliva_pl <- saliva_sum %>%
  ggplot() +
  facet_wrap(~ year, nrow = 1) +
  geom_rect(mapping = aes(ymax = ymax, ymin = ymin, xmax = 4, xmin = 3, fill = virus)) +
  geomtextpath::coord_curvedpolar(theta = "y", start = 1.9*pi) +
  geom_text(mapping = aes(x = 5.25, y=labelPosition, label=label, color=virus), size = 8/cm(1)) +
  geom_text(data = data.frame(x = 1, y = 1, year = c("2021/22", "2022/23"), label = c("2021/22", "2022/23")), mapping = aes(x = x, label = label, y = y), 
            size = 8/cm(1), fontface = 2) +
  scale_x_continuous(limits = c(1, 5.25), expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0),  position = "top") +
  scale_fill_manual(values = wes_palette("IsleofDogs1")) +
  scale_color_manual(values = wes_palette("IsleofDogs1")) +
  ggtitle("a") +
  theme_void() +
  theme(legend.position = "none",
        plot.title = element_text(size = 10, face = 2),
        plot.title.position = "plot",
        legend.key.height = unit(.1, "cm"),
        legend.key.width = unit(.3, "cm"),
        strip.text = element_blank()) +
  guides(fill = guide_legend(nrow = 1),
         color = "none")

saliva_pl
```

## Time distribution

```{r}
saliva_week <- saliva %>%
  mutate(IF = IFA + IFB,
         class = ifelse(year_num == 2022, ifelse(class == "D", "B", "A"), class)) %>%
  dplyr::select(year_num, week, class, intervention, IF, HRV, AdV, CoV, MPV, PIV) %>%
  melt(c("year_num", "week", "class", "intervention")) %>%
  rename(virus = variable) %>%
  group_by(year_num, class, week, intervention, virus) %>%
  summarize(n = sum(value, na.rm = T)) %>%
  ungroup() 

air_week <- air %>%
  mutate(IF = IFA + IFB,
         class = ifelse(year_num == 2022, ifelse(class == "D", "B", "A"), class)) %>%
  dplyr::select(year_num, week, class, intervention, IF, HRV, AdV, CoV, MPV, PIV) %>%
  melt(c("year_num", "week", "class", "intervention")) %>%
  rename(virus = variable) %>%
  group_by(year_num, class, week, intervention, virus) %>%
  summarize(n = sum(value, na.rm = T)) %>%
  ungroup()

saliva_air_week <- rbind(saliva_week %>% mutate(type = "Saliva"),
                         air_week %>% mutate(type = "Bioaerosol")) %>%
  slice(rep(1:n(), times = n)) %>%
  mutate(year = ifelse(year_num == 2022, "2021/22", "2022/23"),
         year = factor(year, levels = c("2021/22", "2022/23")),
         virus = factor(virus, levels = c("IF", "HRV", "AdV", "CoV", "MPV", "PIV")),
         type = factor(type, levels = c("Bioaerosol", "Saliva")),
         type_num = ifelse(type == "Bioaerosol", 1, 2),
         classroom = ifelse(class == "A", "Classroom 1", "Classroom 2"),
         classroom = factor(classroom, levels = c("Classroom 1", "Classroom 2")))

backgrd_y <- saliva %>%
  group_by(school, class, week, year_num) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(
    classroom = ifelse(school == 1, "Classroom 1", ifelse(class == "A", "Classroom 1", "Classroom 2")),
    year = ifelse(year_num == 2022, "2021/22", "2022/23")
  ) %>%
  dplyr::select(year, classroom, week, intervention) %>%
  group_by(year, classroom, week) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(
    intervention = fct_recode(intervention, !!! c("Mask wearing" = "masks", "Air cleaners" = "air cleaners", "No intervention" = "none")),
    intervention = factor(intervention, levels = c("No intervention", "Mask wearing", "Air cleaners")),
    type_num = 1) 

time_pl <- ggplot(mapping = aes(x = week, y = type_num)) +
  facet_grid(classroom ~ year) +
  geom_vline(data = backgrd_y, mapping = aes(xintercept = week + .5), alpha = .1) +
  geom_hline(data = data.frame(y = 1.5), mapping = aes(yintercept = y), linetype = "dashed") +
  geom_point(data = saliva_air_week %>%
               mutate(ps = .17) %>%
               group_by(year, classroom, week, virus, type) %>%
               mutate(type_num = type_num + cumsum(ps) - .3) %>%
               ungroup(),
             mapping = aes(color = virus),
             position = position_dodge(width = .75)) +
  geom_rect(data = backgrd_y,
            mapping = aes(xmin = week - 0.5, xmax = week + 0.5, ymin = -Inf, ymax = Inf, fill = intervention),
            alpha = 0.2) +
  scale_color_manual(values = wes_palette("IsleofDogs1")) +
  scale_fill_manual(values = c("grey", wes_palette("Zissou1")[1], wes_palette("GrandBudapest1")[2])) +
  scale_x_continuous(breaks = 1:7, limits = c(0.5, 7.5), expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 3)) +
  annotate("text", y = 2.8, x = 0.65, label = "Saliva", hjust = 0, size = 8 / cm(1)) +
  annotate("text", y = 0.2, x = 0.65, label = "Bioaerosol", hjust = 0, size = 8 / cm(1)) +
  labs(x = "Study week", title = "b", fill = "Intervention") +
  theme_bw2() +
  theme(axis.title.y = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text = element_text(face = 2, size = 8),
        panel.background = element_rect(fill = "white"),
        plot.background = element_rect(fill = "white"),
        legend.position = "bottom",
        legend.key.width = unit(0.4, "cm"),
        legend.key.height = unit(0.5, "cm"),
        text = element_text(size = 8),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  guides(shape = "none",
         color = "none")

time_pl
```

## Modeling saliva-air pairs

### Checks

Look at the difference in days between positive air and saliva samples. 
* Are air samples detected later or earlier than a positive saliva samples?

```{r}
air_pos <- air %>%
  mutate(IF = IFA + IFB) %>%
  dplyr::select(-IFA, -IFB) %>%
  melt(c("year_num", "school", "class", "week", "date", "sampler", "intervention")) %>%
  filter(!is.na(value), value > 0) %>%
  mutate(class = ifelse(year_num == 2022, ifelse(class == "A/B", "A", "B"), class)) %>%
  dplyr::select(-school, -value, -intervention) %>%
  rename(date_air = date) %>%
  mutate(sample_id = 1:n())

sal_pos <- saliva %>%
  mutate(IF = IFA + IFB) %>%
  dplyr::select(-IFA, -IFB) %>%
  melt(c("year_num", "school", "class", "week", "date", "intervention")) %>%
  filter(!is.na(value), value > 0) %>%
  mutate(class = ifelse(year_num == 2022, ifelse(school == 1, "A", "B"), class)) %>%
  dplyr::select(-school, -value, -intervention) %>%
  rename(date_sal = date)

air_sal_pos <- air_pos %>% 
  dplyr::select(-week) %>%
  left_join(sal_pos %>% dplyr::select(-week)) %>%
  mutate(tdays = difftime(date_air, date_sal, units = "days")) %>%
  group_by(sample_id) %>%
  arrange(abs(tdays)) %>%
  slice(1) %>%
  ungroup()

hist(air_sal_pos$tdays %>% as.numeric, bre = 30)
```

Consider the issue of the missing VIVAS sampler in class B in 2023.
* How many weeks were there when VIVAS detected something but not Coriolis?

```{r}
air_pos_weeks <- air %>%
  mutate(IF = IFA + IFB) %>%
  dplyr::select(-IFA, -IFB) %>%
  melt(c("year_num", "school", "class", "week", "date", "sampler", "intervention")) %>%
  rename(virus = variable) %>%
  mutate(class = ifelse(year_num == 2022, ifelse(class == "A/B", "A", "B"), class)) %>%
  group_by(year_num, class, week, sampler, virus) %>%
  summarise(detected = ifelse(sum(value, na.rm = T) > 0, 1, 0)) %>%
  ungroup() %>%
  group_by(year_num, class, week, virus) %>%
  filter(any(detected == 1)) %>%
  ungroup() %>%
  filter(!(year_num == 2023 & class == "B")) %>%
  arrange(year_num, class, week, virus, sampler) %>%
  dcast(year_num + class + week + virus ~ sampler)

table(air_pos_weeks$Coriolis, air_pos_weeks$Vivas)

3 / 7
```

### Data

```{r}
# saliva bioaerosol pairs weekly
pair_df <- rbind(saliva_week %>% mutate(type = "Saliva"),
                 air_week %>% mutate(type = "Bioaerosol")) %>%
  group_by(year_num, class, virus, week, intervention, type) %>%
  summarize(n = sum(n)) %>%
  ungroup() %>%
  dcast(year_num + class + virus + week + intervention ~ type) %>%
  mutate(is_cov = ifelse(virus == "CoV", "yes", "no"),
         across(c(Bioaerosol, Saliva), ~ ifelse(.x > 0, "yes", "no")),
         across(c(Bioaerosol, Saliva, is_cov), ~ factor(.x, levels = c("no", "yes"))),
         year = ifelse(year_num == 2022, "Winter 2021/2022", "Winter 2022/2023"),
         year = factor(year, levels = c("Winter 2021/2022", "Winter 2022/2023")),
         virus = factor(virus, levels = c("IF", "HRV", "AdV", "CoV", "MPV", "PIV")),
         intervention = factor(intervention, levels = c("none", "masks", "air cleaners"))) 

# add ventilation
vent <- left_join(
  air %>% 
    mutate(date = as.Date(date),
           class = ifelse(class == "A/B", "A", ifelse(class == "D", "B", class))) %>%
    group_by(year_num, class, week, date) %>%
    slice(1) %>%
    ungroup() %>%
    dplyr::select(year_num, class, week, date),
  odf %>%
    group_by(school, classroom, date) %>%
    slice(1) %>%
    ungroup() %>%
    mutate(year_num = lubridate::year(date),
           class = ifelse(classroom == "AB", "A",
                          ifelse(year_num == 2022 & school == "School 2" & classroom == "A", "B", classroom))) %>%
    dplyr::select(year_num, class, date, co2.max) 
) %>%
  group_by(year_num, class, week) %>%
  summarize(maxCO2_imputed = ifelse(is.na(co2.max), T, F), 
            co2.max = mean(co2.max, na.rm = T)) %>%
  ungroup() %>%
  group_by(year_num) %>%
  mutate(co2.max = ifelse(is.na(co2.max), mean(co2.max, na.rm = T), co2.max)) %>%
  ungroup() %>%
  rename(maxCO2 = co2.max) 

pair_df <- left_join(pair_df, vent) 

pair_df %>%
  group_by(year_num, class, week) %>%
  slice(1) %>%
  ungroup() %>%
  filter(!maxCO2_imputed) %>%
  group_by(year_num) %>%
  summarize(mean(maxCO2), sd(maxCO2)) %>%
  mutate_if(is.numeric, round, 0)

# create stan data file
stanDF <- list(
  N = nrow(pair_df),
  K_virus = n_distinct(pair_df$virus),
  bioaerosol = ifelse(pair_df$Bioaerosol == "yes", 1, 0),
  saliva = ifelse(pair_df$Saliva == "yes", 1, 0),
  cov = ifelse(pair_df$is_cov == "yes", 1, 0),
  virus = as.integer(pair_df$virus),
  masks = ifelse(pair_df$intervention == "masks", 1, 0),
  air_cleaners = ifelse(pair_df$intervention == "air cleaners", 1, 0),
  co2max = (pair_df$maxCO2 - mean(pair_df$maxCO2)) / sd(pair_df$maxCO2) 
)
```


### Priors

```{r}
# add prior scales
stanDF$s_saliva <- sd(stanDF$bioaerosol) / sd(stanDF$saliva)
stanDF$s_cov <- sd(stanDF$bioaerosol) / sd(stanDF$cov)
stanDF$s_masks <- .5 * sd(stanDF$bioaerosol) / sd(stanDF$masks)
stanDF$s_air <- .5 * sd(stanDF$bioaerosol) / sd(stanDF$air_cleaners)
stanDF$s_co2 <- sd(stanDF$bioaerosol)

# add prior skew
stanDF$t_masks <- .75
stanDF$t_air <- .75
stanDF$t_co2 <- .25

# add offset
stanDF$offsets = ifelse(pair_df$year_num == 2022, ifelse(pair_df$class == "A", log(2), log(1)), 
                        ifelse(pair_df$class == "A", log(1), log(.5)))

# informative priors
#' positive effect of saliva based on conditional frequencies

table(pair_df$Saliva, pair_df$Bioaerosol) 
# 5/3 odds of positive bioaerosol for positive saliva
stanDF$m_saliva <- sum(pair_df$Saliva[pair_df$Bioaerosol == "yes"] == "yes") / sum(pair_df$Saliva[pair_df$Bioaerosol == "yes"] == "no")

#' 80% prior prob. for a negative effect of interventions
#' smaller effects sizes expected for interventions 

stanDF$m_masks <- optim(par = 0, 
                 fn = function(l, s, k, p) {
                   q <- qasym_laplace(p, l, s, k)
                   abs(q)
                   }, 
                 s = stanDF$s_masks, k = stanDF$t_masks, p = 0.9)$par[1] 
stanDF$m_air <- optim(par = 0, 
                 fn = function(l, s, k, p) {
                   q <- qasym_laplace(p, l, s, k)
                   abs(q)
                   }, 
                 s = stanDF$s_air, k = stanDF$t_air, p = 0.9)$par[1] 

#' same prior for co2 as for intervention effects, but positive

stanDF$m_co2 <- optim(par = 0, 
                 fn = function(l, s, k, p) {
                   q <- qasym_laplace(p, l, s, k)
                   abs(q)
                   }, 
                 s = stanDF$s_co2, k = stanDF$t_co2, p = 0.1)$par[1] 


#' uninformative prior for mean of cov
stanDF$m_cov <- 1
```

```{r}
a  <- seq(-5, 5, .1)
plot(a, dasym_laplace(a, stanDF$m_masks, stanDF$s_masks, stanDF$t_masks))
```


### Model

```{r}
# fit model
mod <- stan(file = "../models/comparison/air-detection-model.stan",
            data = stanDF,
            cores = 4, chains = 4, iter = 2000,
            seed = 12345)

summary(mod, pars = c("alpha", 
                      # "tau_virus", paste0("beta_virus[", 1:stanDF$K_virus, "]"), 
                      "beta_saliva", 
                      "beta_cov", 
                      "beta_masks", 
                      "beta_air_cleaners", "beta_co2max"))$summary %>%
  data.frame() %>%
  add_rownames() %>%
  as_data_frame() %>%
  mutate_if(is.numeric, round, 2)
```

### Results

```{r}
# adjusted OR of mask wearing
mod %>%
  spread_draws(beta_cov) %>%
  mutate(beta_cov = exp(beta_cov)) %>%
  median_qi() %>%
  mutate_if(is.numeric, round, 1)

# posterior probability
post_prob_p <- mod %>%
  spread_draws(beta_cov) %>%
  ungroup() %>%
  summarize(p = round(sum(beta_cov > 1) / n() * 100)) %>%
  dplyr::select(p) %>% unlist()


# predicted probabilities
pred_prob <- mod %>%
  spread_draws(logit_cov, logit_nonCov) %>%
  mutate(CoV = invlogit(logit_cov),
         `non-Cov` = invlogit(logit_nonCov)) %>%
  dplyr::select(-logit_cov, -logit_nonCov) %>%
  melt(c(".chain", ".iteration", ".draw"))

prob_pl <- ggplot(pred_prob, aes(x = variable, y = value)) +
  geom_errorbar(data = pred_prob %>% group_by(variable) %>% median_qi(value),
                mapping = aes(x = variable, ymin = .lower, ymax = .upper, color = variable),
                width = .25) +
  geom_boxplot(data = pred_prob,
               mapping = aes(color = variable),
               position = position_dodge(width = .5), outlier.shape = NA, coef = 0, width = 0.25) +
  stat_summary(data = pred_prob %>% dplyr::select(value, variable),
               mapping = aes(x = variable, y = value, color = variable), geom = "point", fun = "median", 
               position = position_dodge2(width = .5), size = 2, shape = 23, fill = "white") +
  ggsignif::geom_signif(
    annotations = paste0("Pr.=", post_prob_p, "%"), 
    y_position = .55, 
    xmin = 1, xmax = 2, 
    textsize = 8 / cm(1), 
    vjust = -.5, 
    tip_length = .075) +
  scale_y_continuous(expand = c(0,0), limits = c(0, NA), labels = function(x) x * 100) +
  scale_color_manual(values = c(wes_palette("IsleofDogs1")[4], "darkred")) +
  #scale_fill_manual(values = c(wes_palette("IsleofDogs1")[4], "darkred")) +
  labs(y = "Prob. of airborne detection (%)", title = "c") +
  theme_custom() +
  theme(legend.position = "none",
        axis.title.x = element_blank())

prob_pl

# CoV vs non-CoV
pred_prob %>%
  group_by(variable) %>%
  median_qi() %>%
  mutate_if(is.numeric, function(x) round(x * 100) )
```

## Viral Loads

```{r}
vl_saliva <- vl %>%
  filter(type == "saliva") %>%
  mutate(virus = ifelse(virus == "IFA" | virus == "IFB", "IF", virus), 
         virus_type = ifelse(virus == "CoV", "CoV", "non-CoV"),
         virus_type = factor(virus_type, levels = c("CoV", "non-CoV")),
         year = ifelse(year_num == 2022, "2021/22", "2022/23"),
         year = factor(year, levels = c("2021/22", "2022/23")),
         virus = factor(virus, levels = c("IF", "HRV", "AdV", "CoV", "MPV", "PIV"))) 

vl_test <- t.test(
  vl_saliva$ct[vl_saliva$virus_type == "CoV"],
  vl_saliva$ct[vl_saliva$virus_type == "non-CoV"])

vl_saliva_pl <- vl_saliva %>%
  ggplot(aes(x = virus_type, y = ct)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(shape = year, color = virus), width = .25, height = 0) +
  ggsignif::geom_signif(
    annotations = paste0("p=", round(vl_test$p.value, 2)), 
    y_position = 44, 
    xmin = 1, xmax = 2, 
    textsize = 8 / cm(1), 
    vjust = -.5, 
    tip_length = .075) +
  scale_color_manual(values = wes_palette("IsleofDogs1")) +
  scale_shape_manual(values = c(1, 17)) +
  scale_y_continuous(limits = c(NA, 46)) +
  labs(y = "Ct value") +
  theme_bw2() +
  theme(axis.title.x = element_blank(),
        legend.position = "top",
        plot.title.position = "plot",
        legend.title = element_blank()) +
  guides(color = guide_legend(nrow = 2))

save_plot(vl_saliva_pl,
          "../results/comparison-viral-loads.png",
          w = 10, h = 12)
```

## Summary plot

```{r}
design <- "12
33"
comparison <- saliva_pl + prob_pl + time_pl + plot_layout(design = design, widths = c(11, 5), heights = c(6, 11))
ggsave(plot = comparison, file = "../results/comparison.pdf", width = 16 / cm(1), height = 17 / cm(1))
ggsave(plot = comparison, file = "../results/comparison.png", width = 16 / cm(1), height = 17 / cm(1))
```